---
title: "CQL2 development report - Milestone: 'Proof of concept version with examples
  that validate the idea of using the package'"
output:
  pdf_document: default
  html_document:
    df_print: tibble
    theme:
      base_font:
        google: IBM Plex Serif
      code_font:
        google: IBM Plex Mono
fontsize: 8,5pt
indent: yes
vignette: "%\\VignetteIndexEntry{CQL2 development report} %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo = FALSE}
/* add a blue border */
div.mybox {
  border-color: blue;
  border-style: solid;
  padding: 0.5em;
  text-align: center;
}
/* Set to blue bold text inside the box */
div.mybox strong {
  color: blue;
}
```

## Introduction

This document reports partial results of CQL2 development. As stated in the project, the first milestone is the development of a "proof of concept" version, from which we can validate the idea of package implementation. In this report, we show how CQL2 package works and provide a variety of examples of its use. The current version of CQL2 package can be accessed in https://github.com/e-sensing/cql2exp.

## CQL2 filter standard

CQL2 is an OGC standard that enables complex filter expressions on OAFeat3 or STAC web services. CQL2 standard states that expressions can be either represented in JSON or TEXT formats. Our implementation intends to convert native R expressions into CQL2 valid expressions without the need of cumbersome nested lists or dictionaries. Also, with the same filter representation, we can make CQL2 filter requisition either in JSON or TEXT formats.

Let's start with an simple example. In the following code, we have a valid CQL2 expression (in TEXT format) which refers to two properties `vehicle_height` and `bridge_clearance`.

```{verbatim}
vehicle_height > (bridge_clearance-1)
```

This filter expression can be passed in HTTP GET verb implemented by an service to retrieve only those features that satisfies the condition. The same expression can be represented in JSON format, which is more suitable to HTTP POST requests:

```{json}
{
  "op": ">",
  "args": [
    {"property":"vehicle_height"},
    {
      "op": "-",
      "args": [
        {"property":"bridge_clearance"},
        1
      ]
    }
  ]
}
```

Note how properties `vehicle_height` and `bridge_clearance` are represented in this format. They are elements of an object containing a `property` member. Also they go as arguments of operators (in this case `>` and `-` operators). 

## Converting R expressions into CQL2 syntax

```{r setup}
library(cql2exp)
```

In R language, the JSON above could be represented by nested lists which would be somewhat cumbersome to write. To produce valid CQL2 filter expressions we use R abstract syntax tree (AST) from R expressions that can be converted either to TEXT or JSON formats. Let's see the same previous example written in R CQL2: 

```{r}
cql2_text(vehicle_height > (bridge_clearance - 1)) # TEXT format
cql2_json(vehicle_height > (bridge_clearance - 1)) # JSON format
```

In both cases, the same CQL2 object representation is built from the expression using AST of R expression evaluation. Then, the object is converted into TEXT or JSON format.

Sometimes, CQL2 filters in TEXT format is represented in the same way as in R expression. However this should not be always the case, as we can see in some examples provided below.

## Basic CQL2

A Basic CQL2 filter expression can be constructed by logically connecting one or more comparison predicates. 

### Data types and literal values

A literal value is any part of an CQL2 filter expression that is used exactly as it is specified in the expression.

The scalar data types are: `character string`, `number`, `boolean`, `timestamp`, and `date`. 

#### character string

```{R}
cql2_text("Via dell'Avvento")
cql2_json("Via dell'Avvento")
```

#### number

```{R}
cql2_text(3.1415)
cql2_json(-100)
```

#### boolean

```{R}
cql2_text(TRUE)
cql2_json(FALSE)
```

#### timestamp

```{R}
cql2_text(timestamp("1969-07-20T20:17:40Z"))
cql2_json(timestamp("1969-07-20T20:17:40Z"))
```

#### date

```{R}
cql2_text(date("1969-07-20"))
cql2_json(date("1969-07-20"))
```

### Property references

Property of an item can be evaluated in the CQL2 filter expression by its name.

```{R}
cql2_text(windSpeed > 1)
cql2_json(windSpeed > 1)
```

> Note: Properties supported by a service should be listed in endpoint `/collections/{collection_id}/queryables`.

### Standard comparison predicates

A comparison predicate evaluates if two scalar expressions satisfy the specified comparison operator.

The standard comparison operators are: `=`, `!=`, `<`, `>`, `<=`, `>=`, and `IS NULL`.

```{R}
cql2_text(city == "Crato")
cql2_json(city == "Jacareí")
```

```{R}
cql2_text(avg(windSpeed) < 4)
cql2_json(avg(windSpeed) < 4)
```

```{R}
cql2_text(balance - 150.0 > 0)
cql2_json(balance - 150.0 > 0)
```

```{R}
cql2_text(updated >= date('1970-01-01'))
cql2_json(updated >= date('1970-01-01'))
```

#### IS NULL operator

```{R}
cql2_text(!is_null(geometry))
cql2_json(!is_null(geometry))
```

## Common Query Language enhancements

### Advanced comparison operators

A comparison predicate evaluates if two scalar expressions satisfy the specified comparison operator.

Advanced comparison operators are: `LIKE`, `BETWEEN`, and `IN`.

#### LIKE operator

```{R}
cql2_text(name %like% "Smith%")
cql2_json(name %like% "Smith%")
```

#### BETWEEN operator

```{R}
cql2_text(between(depth, 100.0, 150.0))
cql2_json(between(depth, 100.0, 150.0))
```

#### IN operator
```{R}
cql2_text(cityName %in% list('Toronto', 'Frankfurt', 'Tokyo', 'New York'))
cql2_json(cityName %in% list('Toronto', 'Frankfurt', 'Tokyo', 'New York'))
```
```{R}
cql2_text(!category %in% list(1, 2, 3, 4))
cql2_json(!category %in% list(1, 2, 3, 4))
```

### Spatial operators

A spatial predicate evaluates if two spatial expressions satisfy the specified spatial operator.

The supported spatial operators are: `S_INTERSECTS`, `S_EQUALS`, `S_DISJOINT`, `S_TOUCHES`, `S_WITHIN`, `S_OVERLAPS`, `S_CROSSES`, and `S_CONTAINS`. 


```{R, message=FALSE}
poly_sf <- sfheaders::sf_polygon(matrix(c(0,0,0,0,1,1), ncol = 2))
cql2_text(s_intersects(geometry, {{poly_sf}}))
cql2_json(s_intersects(geometry, {{poly_sf}}))
```

> Note: We will use the `sfheaders` package to represent spatial geometries. The `sfheaders` package is backwards compatible with the `sf` package.

> Note: We provide an escape to evaluate user variables using `{{` or `!!`. Both symbols are largely used in R Data Science community.

### Temporal operators

A temporal predicate evaluates if two temporal expressions satisfy the specified temporal operator.

The supported temporal operators are: `T_AFTER`, `T_BEFORE`, `T_CONTAINS`, `T_DISJOINT`, `T_DURING`, `T_EQUALS`, `T_FINISHEDBY`, `T_FINISHES`, `T_INTERSECTS`, `T_MEETS`, `T_METBY`, `T_OVERLAPPEDBY`, `T_OVERLAPS`, `T_STARTEDBY`, and `T_STARTS`.

```{R}
cql2_text(t_intersects(event_date, interval("1969-07-16T05:32:00Z", "1969-07-24T16:50:35Z")))
cql2_json(t_intersects(event_date, interval("1969-07-16T05:32:00Z", "1969-07-24T16:50:35Z")))
```

### Support for functions in CQL2

Functions allow implementations to extend the language.

#### Example of a function that returns a geometry value.

```{R}
cql2_text(s_within(road, Buffer(geometry, 10, "m")))
cql2_json(s_within(road, Buffer(geometry, 10, "m")))
```

> Note: Functions supported by a service should be listed in endpoint `/functions`.

## Concluding remarks

This report describes the delivery of the milestone "Proof of concept version with examples that validate the idea of using the package." 

To complete this milestone, activities A1 and A2 were performed.

We are now in month 2 (M2) of the project. This milestone was scheduled to be delivered in month 3 (M3), after the completion of activity A2. This early delivery will allow us to work with more time on the remaining activities.

The next step of the project is to complete the milestone "Implementing integration testing with STAC services supporting CQL2" which involves activities A3 and A4. This integration will be performed on the Microsoft Planetary computer service.

